Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-13 index.qmd
section:issues; process:write

# Thurstone's theory {#sec-thurstone_theory}
<!-- recording: Sven and Tine 24.11.18; time: 00:04:30 - 00:14:50 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:19:04 - 00:21:37 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:00:35 - 00:06:55 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:30:35 - 00:35:10 -->

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their perception of the stimulus trait. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), called the *modal discriminal process*, indicates the stimulus position on the trait continuum, while its dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates the discriminal processes along a quality trait continuum for two written texts. The figure shows that these processes individually follow a Normal distribution. Furthermore, it shows that the modal discriminal process for Text B is positioned further along the continuum than Text A $(S_{B} > S_{A})$. Additionally, the figure highlights the broader distribution of Text B compared to Text A as a result of its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

![Discriminal processes for two written texts](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law states that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait. This law emphasizes that the relative distance between stimuli, rather than their absolute positions on the continuum, determines the outcome of the pairwise comparison. Indeed, the theory assumes that the observed dichotomous outcome arises from the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*. Since the individual discriminal processes follow a Normal distribution, the discriminal difference also follows a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli and the dispersion captures the variability of that relative separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the texts presented in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion is evident from the positive difference in the modal discriminal processes of the texts $(S_{B} - S_{A} > 0)$ and the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$ (shaded gray area). As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

![Discriminal difference for two written texts](/images/figures/discriminal_difference.png){#fig-discriminal_difference width=70%}

@@

<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:00:25 - 00:04:30 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:14:50 - 00:19:04 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:06:55 - 00:24:50 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->

<!-- 1. Thurstone's Cases and their simplifying assumptions  -->
The previous section outlines the general form of Thurstone's theory, which applies to a CJ design where a single judge evaluates multiple stimuli. For the practical application of the theory, Thurstone developed four additional cases derived from this general form, where each successive case incorporates additional simplifying assumptions. Case I represents the general form of the theory. Case II extends this by allowing multiple judges to make comparisons rather than restricting the comparisons to a single judge. Case III introduces the assumption of zero correlation between stimuli. Case IV builds on this by assuming the stimuli have similar dispersions. Finally, Case V replaces this assumption with the condition that the stimuli have equal discriminal dispersions. @tbl-thurstone_cases summarizes these cases and their assumptions. For a detailed discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

<!-- , suggesting all stimuli have the same variability in their perceived trait levels -->

![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#tbl-thurstone_cases width=90%}

<!-- 2. The Case V and its simpler statistical representation: the BTL model -->
Despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (see @fig-logistic_vs_normal).

@@

::: {#fig-logistic_vs_normal layout-nrow=2}

![Probability density](/images/figures/density.png){#fig-density width=70%}

![Cummulative probability](/images/figures/cummulative.png){#fig-cumulative width=70%}

Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255].
:::

<!-- 3. But what are limitations of Case V (and the BTL model)? -->
Nevertheless, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions might be problematic when complex traits or heterogeneous stimuli are involved [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
The discriminal dispersions of stimuli are crucial in determining the dichotomous outcomes of pairwise comparisons. Discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-dispersion illustrates how more uncertainty in the trait perception of one text relative to another broadens the distribution of their discriminal difference. This broadening affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B > A)$. Additionally, the figure shows that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor one text over the other than when their dispersions differ (shaded gray area).

![Discrepancies in the stimuli dispersions and tehir effect on the distribution of the discriminal difference](/images/figures/dispersion.png){#fig-dispersion width=70%}

<!-- 2. But does it hold?  -->
However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics. These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than others [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. This finding implies that discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. Moreover, the presence of these "misfit" judges and their deviations can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue1b.

<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs. [@Thurstone_1927a, pp. 376] -->

@@

<!-- 3. But what are the challenges when these assumptions are violated? -->
Therefore, assuming equal dispersions between stimuli, despite its violation, can cause Case V (and the BTL model) to inflate the reliability of the comparison outcome. Moreover, ignoring the difference in dispersions can lead to the neglect of critical differences in the reliability of the trait across stimuli, resulting in inaccurate conclusions about the trait's estimates [@McElreath_2020, pp. 370]. Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are unpredictable, as they depend on the specific stimuli researchers exclude from the analysis. Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Similar to the discriminal dispersions, the correlation between discriminal processes plays a crucial role in determining the dichotomous outcomes of pairwise comparisons. This correlation also shapes the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-correlation illustrates how the dependence in trait perception between two texts narrows the distribution of their discriminal difference. This narrowing affects the area under the curve where the discriminal difference distinctly favors one text over the other, $P(B>A)$. Moreover, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor one text over the other than when the texts are highly correlated (shaded gray area).

@@

![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/figures/correlation.png){#fig-correlation width=70%}

<!-- 2. But does it hold?  -->
Notably, Thurstone's Case V and the BTL model assume independent trait perceptions across stimuli. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, can introduce additional statistical and measurement issues, as discussed in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, the psychometric and statistical literature emphasizes the need to address factors that create dependencies between stimuli, as failing to do so can affect the reliability of the comparison outcomes and lead to inaccurate conclusions about the trait's estimates. For instance, researchers who overlook additional relevant traits, such as judges' biases, can cause dimensional mismatches in the statistical model used for analysis. This mismatch can artificially inflate the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989]. Furthermore, as discussed in @sec-theory-issue1a, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of reliability [@Hoyle_et_al_2023, pp. 482]. Together, these issues undermine the reliability of the trait and compromise the validity of its estimates [@Perron_et_al_2015, pp. 2].

@@

@@

@@

<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independece assumption -->

>>>>

# Thurstone's theory {#sec-thurstone_theory}
<!-- recording: Sven and Tine 24.11.18; time: 00:04:30 - 00:14:50 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:19:04 - 00:21:37 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:00:35 - 00:06:55 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:30:35 - 00:35:10 -->
<!-- recording: Sven and Tine 24.12.09; time: 00:00:00 - 00:03:20 -->

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons where a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

<!-- The psychological continuum or scale is so constructed or defined that the frequencies of the respective discriminal processes for any given stimulus form a normal distribution on the psychological scale. This involves no assumption of a normal distribution or of anything else. The psychological scale is at best an artificial construct. If it has any physical reality, we certainly have not the remotest idea of what it may be like. ... We define the scale in terms of the frequencies of the discriminal processes for any stimulus. This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, pp. 266] -->

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

![Discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=70%}

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces *the law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as *the discriminal difference*, determines the observed dichotomous outcome. Moreover, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

![Discriminal difference for two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=70%}

@@

<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:00:25 - 00:04:30 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:14:50 - 00:19:04 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:06:55 - 00:24:50 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->
<!-- recording: Sven and Tine 24.12.09; time: 00:03:20 - 00:08:10 -->

<!-- 1. What is the problem with the general form -->
Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form. Each case progressively incorporated additional simplifying assumptions into the model.

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
In Case I, Thurstone assumed that pairs of stimuli maintained a constant correlation across all comparisons. In Case II, he allowed multiple judges to make comparisons instead of restricting evaluations to a single judge. In Case III, he introduced the assumption of zero correlation between stimuli. In Case IV, he assumed stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For an in-depth discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248–253].

<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->

![Thurstones cases and their asumptions](/images/png/thurstone_cases.png){#tbl-thurstone_cases width=100%}

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- ::: {#fig-logistic_vs_normal layout-nrow=2} -->

<!-- ![Probability density](/images/png/density.png){#fig-density width=70%} -->

<!-- ![Cummulative probability](/images/png/cummulative.png){#fig-cumulative width=70%} -->

<!-- Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255]. -->
<!-- ::: -->

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- recording: Sven and Tine 24.12.09; time: 00:08:10 - 00:37:00 -->

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, the discriminal dispersions of stimuli play a critical role in determining the outcome of pairwise comparisons. Specifically, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-dispersion illustrates this idea, assuming a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersion for Text A remains constant and that the texts have no correlation $(\rho=0)$.

@fig-dispersion reveals that more uncertainty in the trait perception of Text B compared to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.

![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/png/dispersion.png){#fig-dispersion width=70%}

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their discriminal processes [@Thurstone_1927a, pp. 373]. For example, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-dispersion when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The estimation accuracy of this discriminal difference, in turn, ensures reliable discriminal process estimates for the texts (citation needed?). It then intuitively follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.

<!-- In experimental practice the procedure is the reverse of this hypothesis because the frequencies are known first experimentally, and from these frequencies we construct the psychological continuum. [@Thurstone_1927a, pp. 373] -->

<!-- 3. But does it hold?  -->
However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of this violation may already exist in the CJ literature as misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts. A similar reasoning applies to "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. Moreover, these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.

<!-- It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series, such as handwriting specimens, English compositions, sewing samples, oriental rugs. In measurements of the type known as judgment scales the discriminal dispersion on the psychological continuum becomes one of the unknowns to be determined as well as the scale value of the specimen. Every specimen in such a series presents two unknown values to be determined, namely, the scale value of its modal discriminal process on the psychological continuum and its discriminal dispersion. [@Thurstone_1927a, pp. 376] -->

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Then, incorrectly assuming equal dispersions between stimuli can lead the BTL model to introduce various statistical and measurement issues. For example, the model could overestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli. This overestimation may result in spurious inferences about these differences [@McElreath_2020, pp. 370] and, by extension, about the stimuli's discriminal processes. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the common practice in CJ literature of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] may inadvertently discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. These biases are often unpredictable, as they depend on the specific stimuli excluded from the analysis.

<!-- But ignoring over-dispersion can also lead to all of the same problems as ignoring any predictor variable. Heterogeneity in counts can be a confound, hiding effects of interest or producing spurious inferences. [@McElreath_2020, pp. 370] -->

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another. Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons. @fig-correlation illustrates this concept, assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersions for both texts remain constant.

@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area), compared to scenarios when the texts are highly correlated.

@@

![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/png/correlation.png){#fig-correlation width=70%}

<!-- 2. But this is not the scenario we observe -->
However, in experimental practice, researchers typically follow this process in reverse. For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts (citation needed?).

<!-- It then follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the assumption of zero correlation. -->

<!-- 2. But does it hold?  -->
Notably, Thurstone's Case V and the BTL model assume independent discriminal processes across comparisons. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
[In any case, similar to @sec-theory-issue1a, incorrectly assuming zero correlation between stimuli can lead the BTL model to introduce various statistical and measurement issues. For instance, the model could over- or underestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli. This over- underestimation may result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes [@Hoyle_et_al_2023, pp. 341]. @fig-correlation also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$. This missaligment can be due to the overlook of additional relevant traits, such as judges' biases, which cause dimensional mismatches in the BTL model, artificially inflating the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989]. Furthermore, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].]{style="color:red;"}

@@

[In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait estimates and ultimately compromise its validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.]{style="color:red;"}

@@

The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons

![Theoretical model A1$](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=29%}

The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference

![Theoretical model A2$](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=40%}

The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes

![Theoretical model A3$](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=60%}

without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in

<!-- Notably, the theory assumes the judge compares each pair of stimuli a "sufficient number of times" to accurately calculate the relative separation probability [@Thurstone_1927b, pp. 267]. -->

<!-- It is assumed that the single observer compares each pair of stimuli sufficient number of times so that a proportion, P(A>B), may be determined for each pair of stimuli [@Thurstone_1927b, pp. 267] -->

<!-- Using assumptions from IRT [@deAyala_2009, p. 20-21], we are assuming a multidimensional model, that violates the local independence assumption -->

@@
